---
title: "Hands-on Exercise 4"
subtitle: "Spatial Weights and Applications"
author: "Le Truong Chinh"
date: "16 September 2026"
date-modified: "17 September 2026"
description: "An executed ISSS626 exercise building spatial weights and spatially lagged variables for Hunan counties."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 3
number-sections: true
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 5.5
---

## Overview

::: {.exercise-lead}
This exercise builds several ways of defining neighbouring counties in Hunan. I then use these relationships to compare each county's GDP per capita with the values around it.
:::

::: {.callout-note title="A few additions to the tutorial 🙂"}
I keep the [Chapter 8](https://r4gdsa.netlify.app/chap08.html) code unchanged and add:

- **Short observations** after the main results.
- **A comparison table and difference map** showing which counties stand out from their neighbours.
:::

## The Study Area and Data

The analysis uses two datasets:

- the Hunan county boundary layer, containing 88 county polygons; and
- `Hunan_2012.csv`, containing selected development indicators for 2012.

### Getting Started

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## Getting the Data Into R Environment

### Import shapefile into r environment

`st_read()` imports the Hunan shapefile as an `sf` object.

```{r}
hunan <- st_read(dsn = "chap08/data/geospatial", 
                 layer = "Hunan")
```

### Import csv file into r environment

```{r}
hunan2012 <- read_csv("chap08/data/aspatial/Hunan_2012.csv")
```

### Performing relational join

The CSV attributes are joined to the county polygons. I retain the fields needed for the rest of the exercise.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

## Visualising Regional Development Indicator

The first map labels the counties, while the second shows the spatial distribution of GDP per capita (`GDPPC`).

```{r fig.width=12, fig.height=8}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

The map gives a first indication that GDPPC is not evenly distributed. Some neighbouring counties have similar values, so the next step is to define which counties count as neighbours.

## Computing Contiguity Spatial Weights

Contiguity weights use shared polygon boundaries. Queen contiguity accepts a shared edge or corner, while Rook contiguity requires a shared edge.

### Computing (QUEEN) contiguity based neighbours

```{r}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

The summary shows how many neighbours each county has and whether any county is disconnected from the rest.

The neighbour IDs for the first county are:

```{r}
wm_q[[1]]
```

The first county is:

```{r}
hunan$County[1]
```

Its five Queen neighbours are:

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

Their GDPPC values are:

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

The full neighbour-list structure can also be inspected directly.

```{r eval=TRUE, echo=TRUE}
str(wm_q)
```

### Creating (ROOK) contiguity based neighbours

```{r}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

Rook contiguity is slightly stricter because a corner alone is not enough to make two counties neighbours.

### Visualising contiguity weights

The neighbour links are drawn between representative points. I extract the longitude and latitude of each polygon centroid and combine them into one coordinate matrix.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

```{r}
coords <- cbind(longitude, latitude)
```

```{r}
head(coords)
```

#### Plotting Queen contiguity based neighbours map

```{r fig.width=8, fig.height=6}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

#### Plotting Rook contiguity based neighbours map

```{r fig.width=8, fig.height=6}
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Plotting both Queen and Rook contiguity based neighbours maps

```{r fig.width=12, fig.height=8}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

The two networks are very similar, but the Queen network includes a few extra links where polygons meet only at a corner.

## Computing distance based neighbours

Distance-based weights do not depend on shared borders. They are useful when nearby areas may influence one another even without touching.

### Determine the cut-off distance

I first find the nearest neighbour of every county and inspect the longest of these links.

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The maximum nearest-neighbour distance is about 61.79 km. A 62 km cut-off therefore gives every county at least one neighbour.

### Computing fixed distance weight matrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

An average of about 3.68 links means that a county has around four neighbours within 62 km, although the actual count differs across counties.

```{r}
str(wm_d62)
```

```{r}
table(hunan$County, card(wm_d62))
```

I check whether the neighbour network is split into separate groups.

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

The result confirms that all counties belong to one connected network.

#### Plotting fixed distance weight matrix

```{r fig.width=8, fig.height=6}
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

The black lines connect counties within 62 km, while the red lines show each county's first nearest neighbour.

```{r fig.width=12, fig.height=8}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

### Computing adaptive distance weight matrix

A six-nearest-neighbour network gives each county the same number of neighbours, even where counties are more widely spaced.

```{r}
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

```{r}
str(knn6)
```

#### Plotting distance based neighbours

```{r fig.width=8, fig.height=6}
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

The fixed-distance method keeps one common distance rule but produces unequal neighbour counts. The six-nearest-neighbour method does the opposite: it keeps the count equal but allows the linking distance to vary.

## Weights based on IDW

Inverse-distance weights give more influence to closer neighbours. I calculate the Queen-neighbour distances and take the inverse of each distance.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

## Row-standardised Weights Matrix

Row standardisation makes the weights for each county sum to one. With equal weights, every neighbour contributes the same share of that county's surrounding value.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

The following output shows the weights for region 10. It has eight neighbours, so each receives a weight of 0.125.

```{r}
rswm_q$weights[10]
```

I also create inverse-distance weights from the Queen neighbour list.

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

These weights vary because a closer pair of counties receives a larger value than a more distant pair.

## Application of Spatial Weight Matrix

### Spatial lag with row-standardized weights

A spatial lag is the weighted value around each county. With row-standardised Queen weights, it is the average GDPPC of the neighbouring counties.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

The first county's neighbour values can be checked again:

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

Their mean is the first value in `GDPPC.lag`. This makes the meaning of the row-standardised spatial lag direct: it summarises the surrounding counties rather than the county itself.

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

```{r}
head(hunan)
```

```{r fig.width=12, fig.height=8}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

The lag map is smoother than the original GDPPC map because every mapped value averages several neighbouring counties. It shows the local setting around each county, not a replacement for the county's own GDPPC.

### Spatial lag as a sum of neighboring values

Binary weights assign a value of 1 to every neighbour. The resulting lag is the total GDPPC of all neighbouring counties.

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
lag_sum
```

Unlike the average, the sum usually becomes larger when a county has more neighbours. It therefore reflects both neighbouring GDPPC and neighbour count.

```{r}
hunan <- left_join(hunan, lag.res)
```

```{r fig.width=12, fig.height=8}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### Spatial window average

The spatial window includes the county itself together with its neighbours. I first add each county to its own neighbour list.

```{r}
wm_qs <- include.self(wm_q)
```

```{r}
wm_qs[[1]]
```

County 1 now has six entries: itself and its five Queen neighbours.

```{r}
wm_qs <- nb2listw(wm_qs)
wm_qs
```

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

```{r}
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

```{r}
hunan <- left_join(hunan, lag_wm_qs.res)
```

```{r}
hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()
```

The two averages differ because `lag GDPPC` uses only neighbouring counties, while `lag_window_avg GDPPC` also includes the county's own GDPPC.

```{r fig.width=12, fig.height=8}
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

### Spatial window sum

The window sum also includes the county itself, but it uses binary rather than row-standardised weights.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

```{r}
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

```{r}
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

```{r}
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()
```

`lag_sum GDPPC` totals the neighbours only. `w_sum GDPPC` adds the county's own GDPPC to that total.

```{r fig.width=12, fig.height=8}
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```

## Some other observations

### Local GDPPC differences

I subtract the neighbour average from each county's own GDPPC. A positive difference means that the county is above its neighbours; a negative difference means that it is below them.

```{r}
hunan <- hunan %>%
  mutate(`GDPPC difference` = GDPPC - `lag GDPPC`)

hunan %>%
  st_drop_geometry() %>%
  select(County, GDPPC,
         `Neighbour average` = `lag GDPPC`,
         Difference = `GDPPC difference`) %>%
  arrange(desc(abs(Difference))) %>%
  slice_head(n = 8) %>%
  mutate(across(where(is.numeric), ~round(.x, 0))) %>%
  kable()
```

```{r fig.width=9, fig.height=6}
ggplot(hunan) +
  geom_sf(aes(fill = `GDPPC difference`),
          colour = "white", linewidth = 0.15) +
  scale_fill_gradient2(
    low = "#2c7bb6", mid = "#f7f7f7", high = "#d7191c",
    midpoint = 0,
    name = "Difference") +
  labs(
    title = "County GDPPC compared with its neighbours",
    subtitle = "Positive values are above the Queen-neighbour average") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right")
```

Lengshuijiang, Changsha and Zixing stand well above their neighbour averages. Pingjiang and Guidong show the largest differences in the other direction. These are useful local contrasts to investigate, but the differences alone do not explain why the counties stand out.

## Takeaways

The choice of spatial weights also changes what “nearby” means. Queen and Rook weights follow shared boundaries, a fixed-distance band uses one common reach, and k-nearest neighbours gives every county the same neighbour count.

## References

- [Chapter 8: Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html)
- [Creating Neighbours using sf objects](https://cran.r-project.org/web/packages/spdep/vignettes/nb_sf.html)
