---
title: "Hands-on Exercise 5"
subtitle: "Global and Local Measures of Spatial Autocorrelation"
author: "Le Truong Chinh"
date: "24 September 2026"
date-modified: "24 September 2026"
description: "An executed ISSS626 exercise examining global and local spatial autocorrelation in Hunan's GDP per capita."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 4
number-sections: true
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 5.5
---

::: {.exercise-lead}
This exercise asks two related questions about GDP per capita in Hunan. First, is there an overall spatial pattern? Second, where are the clusters and unusual counties?
:::

::: {.callout-note title="A small addition to the tutorial 🙂"}
I keep the code from [Chapter 9](https://r4gdsa.netlify.app/chap09.html) and [Chapter 10](https://r4gdsa.netlify.app/chap10.html) unchanged. I add short explanations after the main results and one summary table at the end.
:::

## Global Measures of Spatial Autocorrelation

### Overview

Global measures give one result for the whole study area. Here, they help me check whether counties with similar GDP per capita tend to be near one another.

### Getting Started

#### The analytical question

Is GDP per capita spread evenly across Hunan, or do similar values form a spatial pattern?

#### The Study Area and Data

The analysis uses the Hunan county boundaries and the 2012 development indicators. There are 88 counties in the boundary layer.

#### Setting the Analytical Toolls

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

### Getting the Data Into R Environment

#### Import shapefile into r environment

```{r}
hunan <- st_read(dsn = "chap09/data/geospatial", 
                 layer = "Hunan")
```

#### Import csv file into r environment

```{r}
hunan2012 <- read_csv("chap09/data/aspatial/Hunan_2012.csv")
```

#### Performing relational join

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

#### Visualising Regional Development Indicator

The same data can look different depending on how values are grouped. Equal intervals use ranges of the same width, while quantiles place a similar number of counties in each group.

```{r}
#| fig-width: 12
#| fig-height: 8
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

The broad pattern is similar in both maps, although the class boundaries change which counties appear in the darkest groups.

### Global Measures of Spatial Autocorrelation

#### Computing Contiguity Spatial Weights

Queen contiguity treats counties that share an edge or a corner as neighbours.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

#### Row-standardised weights matrix

Row-standardised weights make the weights around each county sum to one.

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

#### What can we learn from the code chunk above?

The neighbour list contains 88 counties. Each neighbouring county receives an equal share of the weight for the county being evaluated.

### Global Measures of Spatial Autocorrelation: Moran's I

#### Maron's I test

Moran's I is positive when counties with similar values tend to be close together.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

The positive Moran's I and very small p-value indicate that GDPPC is spatially clustered rather than randomly arranged.

#### Computing Monte Carlo Moran's I

The permutation test compares the observed result with 999 random arrangements of the same values.

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

The observed result is more extreme than all the simulated values, which supports the same conclusion.

#### Visualising Monte Carlo Moran's I

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

Most random results sit near zero, while the observed value is clearly higher.

### Global Measures of Spatial Autocorrelation: Geary's C

#### Geary's C test

Geary's C focuses more on differences between neighbouring counties. A value below one indicates that neighbours tend to be similar.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

The result is below one with a small p-value, so it agrees with Moran's I.

#### Computing Monte Carlo Geary's C

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

#### Visualising the Monte Carlo Geary's C

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

The random results are centred close to one. The observed value is lower, showing that neighbouring counties are more alike than expected by chance.

### Spatial Correlogram

A correlogram checks whether the pattern remains as the number of neighbour steps increases.

#### Compute Moran's I correlogram

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

```{r}
print(MI_corr)
```

The positive relationship is strongest for nearby counties and weakens with distance. At later neighbour steps, the sign becomes negative.

#### Compute Geary's C correlogram and plot

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```

The Geary result tells a similar story: the clearest similarity is among the closest counties.

## Local Measures of Spatial Autocorrelation

### Overview

The global tests confirm that a pattern exists. Local measures show where the pattern occurs and separate clusters from spatial outliers.

### Getting Started

#### The analytical question

Which counties form clusters of high or low GDPPC, and which counties differ from their surroundings?

#### The Study Area and Data

The same Hunan datasets are used again. This part projects the boundary to UTM Zone 50N before calculating distances.

#### Setting the Analytical Toolls

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

### Getting the Data Into R Environment

#### Import shapefile into r environment

```{r}
hunan <- st_read(dsn = "chap10/data/geospatial", 
                 layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

#### Import csv file into r environment

```{r}
hunan2012 <- read_csv("chap10/data/aspatial/Hunan_2012.csv")
```

#### Performing relational join

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

#### Visualising Regional Development Indicator

```{r}
#| fig-width: 12
#| fig-height: 8
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

### Local Indicators of Spatial Association(LISA)

#### Computing Contiguity Spatial Weights

```{r}
wm_q <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1)
```

```{r}
summary(wm_q$nb)
```

#### Row-standardised weights matrix

The `wt` column stores the row-standardised weights for each county's neighbours.

#### What can we learn from the code chunk above?

These weights let every county be compared fairly with its own set of neighbours, even when the neighbour counts differ.

#### Computing local Moran's I

```{r}
set.seed(1234)
```

```{r}
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

```{r}
glimpse(lisa)
```

##### Mapping the local Moran's I

##### Mapping local Moran's I values

```{r}
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Morans'I",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Morans'I of GDPPC (Queen's method)")
```

##### Mapping local Moran's I p-values

```{r}
tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")
```

##### Mapping both local Moran's I values and p-values

```{r fig.width=12, fig.height=8}
ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Moran's I of GDPPC (Queen's method)")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")

tmap_arrange(ii.map, p_ii.map, asp=1, ncol=2)
```

The p-value map is important because a large local Moran's I value is not automatically a reliable cluster.

#### Preparing and Visualising LISA Map

##### Plotting Moran scatterplot

```{r}
lisa <- lisa %>%
  mutate(lag_GDPPC = st_lag(
    GDPPC, nb, wt),
    .before = 1) %>%
  unnest(lag_GDPPC)
```

```{r}
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "red") +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot") +
  theme_minimal()
```

```{r}
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$lag_GDPPC), lty=2) + 
    geom_vline(xintercept=mean(lisa$GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

The four groups separate counties that match their surroundings from counties that stand out locally.

##### Plotting Moran scatterplot with standardised variable

```{r}
lisa <- lisa %>%
  mutate(z_GDPPC = scale(GDPPC),
         z_lag_GDPPC = scale(lag_GDPPC),
         .before = 1)
```

```{r}
ggplot(data = lisa, 
       aes(x = z_GDPPC, 
           y = z_lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$z_lag_GDPPC), lty=2) + 
    geom_vline(xintercept=mean(lisa$z_GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised Spatial Lag of GDPPC",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

Standardising changes the scale but not which quadrant each county belongs to.

##### Preparing LISA map classes

```{r}
signif <- 0.05       
```

##### Plotting and visualising LISA map

```{r}
lisa <- lisa %>%
  mutate(
    LISA_cluster = ifelse(
      p_ii < 0.05,
      as.character(mean), 
      "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

```{r}
lisa_map <- tm_shape(lisa) + 
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "blue",        # Low-Low
        "lightblue",   # Low-High
        "pink",        # High-Low
        "red"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8

tmap_arrange(ii.map, lisa_map, asp=1, ncol=2)
```

The cluster map keeps only counties with p-values below 0.05. Grey counties do not have enough evidence to assign a cluster type.

### Hot Spots and Cold Spots Analysis (HCSA)

Gi* looks for groups of unusually high or low values within a distance-based neighbourhood.

#### Deriving distance weight matrix

##### Computing fixed distance weights

```{r}
ct <- critical_threshold(st_geometry(hunan))
ct
```

The threshold is the shortest distance that ensures every county has at least one neighbour.

```{r}
hunan_fdw <- hunan %>% 
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry), 
                 upper = ct)),
    wt = st_weights(nb, 
                      style = "W"),
    .before = 1) 
```

##### Computing adaptive distance weights

```{r}
hunan_adw <- hunan %>%
  mutate(nb = include_self(
    st_knn(
      st_geometry(geometry),
      k = 6)),
    wt = st_weights(
      nb, style = "W"),
    .before = 1)
```

The fixed-distance method keeps one distance for all counties. The adaptive method instead keeps six neighbours for each county.

##### Computing adaptive distance weights

#### Computing local Gi\*

```{r}
HCSA_fdw <- hunan_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```

#### Preparing and Visualising HCSA Map

##### Mapping Gi\* with fix distance weights

```{r}
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 6,
                values = "brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "Gi*",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Gi* of GDPPC (Fixed Bandwidth d = 60799.91m)")
```

##### Mapping local Gi\* p-values

```{r}
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "simulated p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (Fixed distance)")
```

##### Mapping both Gi\* values and p-values

```{r fig.width=12, fig.height=8}
Gi_star_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p_values_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (fixed distance)")

tmap_arrange(Gi_star_map, p_values_map, asp=1, ncol=2)
```

##### Plotting and visualising HCSA Map

```{r}
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low"  ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )
```

```{r}
HCSA_map <- tm_shape(HCSA_fdw) + 
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "red",        # Low-Low
        "blue"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "HSCA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("HCSA Clusters (p < 0.05)")
HCSA_map
```

```{r}
#| fig-width: 12
#| fig-height: 8

tmap_arrange(Gi_star_map, HCSA_map, asp=1, ncol=2)
```

The final map separates significant hot and cold areas from counties without a clear result.

## Some other observations

I place the two local classifications into one small table. This makes it easier to see how many counties receive a clear label under each method.

```{r}
bind_rows(
  lisa %>%
    st_drop_geometry() %>%
    count(Category = LISA_cluster) %>%
    mutate(Method = "Local Moran's I"),
  HCSA_fdw %>%
    st_drop_geometry() %>%
    count(Category = HCSA_cluster) %>%
    mutate(Method = "Gi*")
) %>%
  select(Method, Category, Counties = n) %>%
  knitr::kable()
```

Local Moran's I identifies 13 counties: eight High-High, three Low-Low and two Low-High. Gi* identifies seven hot spots and three cold spots. The counts differ because the two methods look for different kinds of spatial patterns.

## Takeaways

Both global tests show that Hunan's GDPPC has a real spatial pattern. The local results then reveal that this pattern is not spread evenly: some counties form clear high or low groups, while others do not have enough evidence for a local label. These results describe where values cluster, but they do not explain the causes.

## References

- [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09.html)
- [Chapter 10: Local Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10.html)
