---
title: "Hands-on Exercise 2"
subtitle: "First- and second-order spatial point pattern analysis"
author: "Le Truong Chinh"
date: "3 September 2026"
description: "An executed ISSS626 exercise analysing Singapore childcare centre locations with spatstat."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 3
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 5.5
---

::: {.exercise-lead}
This exercise follows Chapters 4 and 5 of *R for Geospatial Data Science and Analytics*. Each tutorial code cell is kept separate, executed, and shown with its result.
:::

## Part 1: First-order spatial point pattern analysis

First-order analysis describes where childcare services are concentrated. I begin by preparing the data, then use the Clark--Evans test and kernel density estimation (KDE).

### 1.1 Packages

```{r}
pacman::p_load(sf, terra, spatstat, 
               tmap, rvest, tidyverse)
```

### 1.2 Importing and preparing the geospatial data

The analysis uses the Master Plan 2019 subzone boundaries and Child Care Services data from data.gov.sg. The current childcare download is GeoJSON, so I prepared local KML and RDS copies with the same records and filenames expected by the tutorial.

```{r}
mpsz_sf <- st_read("chap04/data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```

The KML stores the boundary attributes in an HTML table. This function extracts one named field from that table.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

The tutorial removes three offshore areas before creating the Singapore study window.

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
write_rds(mpsz_cl, 
          "chap04/data/mpsz_cl.rds")
```

```{r}
mpsz_cl <- read_rds("chap04/data/mpsz_cl.rds")
```

The childcare locations are also transformed to Singapore SVY21 (EPSG:3414), whose units are metres.

```{r}
childcare_sf <- st_read("chap04/data/ChildCareServices.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

### 1.3 Mapping the two datasets

The first map checks that the childcare points and subzone boundaries share the same spatial extent.

```{r}
tm_shape(mpsz_cl) +
  tm_polygons() +
tm_shape(childcare_sf)+
  tm_dots()
```

The interactive version makes it possible to zoom and inspect individual points. The final line returns `tmap` to static mode.

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
tmap_mode('plot')
```

### 1.4 Converting the data for `spatstat`

`spatstat` represents points as a `ppp` object and the study area as an `owin` object.

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

```{r}
class(childcare_ppp)
```

```{r}
summary(childcare_ppp)
```

```{r}
sg_owin <- as.owin(mpsz_cl)
```

```{r}
class(sg_owin)
```

```{r}
plot(sg_owin)
```

Only childcare points inside the Singapore observation window are retained.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

```{r}
childcareSG_ppp
```

### 1.5 Clark--Evans nearest-neighbour test

The Clark--Evans test compares observed nearest-neighbour spacing with complete spatial randomness (CSR). An R value below 1 indicates that points are closer together than expected.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"))
```

The Monte Carlo version compares the observed pattern with 99 simulated random patterns.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

Both tests provide evidence that childcare services are clustered rather than randomly distributed.

### 1.6 Kernel density estimation

KDE turns the individual service locations into a continuous density surface. The first estimate uses Diggle's automatic bandwidth and an edge correction.

```{r}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 
```

```{r}
#| fig-height: 6
plot(kde_SG_diggle)
```

```{r}
summary(kde_SG_diggle)
```

The very small values are services per square metre. The selected bandwidth can be inspected directly.

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

Rescaling the point pattern from metres to kilometres makes the density values easier to read.

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp, 1000, "km")
```

```{r}
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma=bw.diggle,
                              edge=TRUE,
                              kernel="gaussian")
```

```{r}
#| fig-height: 6
plot(kde_childcareSG_km)
```

The map pattern is unchanged, but the values are now expressed per square kilometre.

#### Comparing automatic bandwidth methods

The following cells calculate the four bandwidth selectors used in the tutorial.

```{r}
bw.CvL(childcareSG_ppp_km)
```

```{r}
bw.scott(childcareSG_ppp_km)
```

```{r}
bw.ppl(childcareSG_ppp_km)
```

```{r}
bw.diggle(childcareSG_ppp_km)
```

A smaller bandwidth preserves more local detail, while a larger bandwidth produces a smoother regional pattern.

```{r}
#| fig-width: 16
#| fig-height: 6
kde_childcareSG.ppl <- density(childcareSG_ppp_km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

#### Comparing kernel shapes

With the same bandwidth, the four kernels distribute weight around each location differently.

```{r fig.width=16, fig.height=12}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

### 1.7 Fixed and adaptive KDE

The fixed estimate applies a 0.6 km bandwidth everywhere.

```{r}
#| fig-height: 6
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                              sigma=0.6, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_childcareSG_fb)
```

The adaptive estimate changes its smoothing according to the local point pattern.

```{r}
#| fig-height: 6
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km, 
  method="kernel")
plot(kde_childcareSG_ab)
```

```{r fig.width=16, fig.height=6}
par(mfrow=c(1,2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```

### 1.8 Converting the KDE surface to a raster map

The `im` density surface is converted to a `SpatRaster` so it can be mapped with `tmap`.

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)
```

```{r}
class(kde_childcareSG_bw_terra)
```

```{r}
kde_childcareSG_bw_terra
```

The raster does not carry a coordinate reference system automatically, so the tutorial assigns EPSG:3414.

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
```

```{r}
kde_childcareSG_bw_terra
```

```{r}
tm_shape(kde_childcareSG_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
            title = "Density values",
            title.size = 0.7,
            text.size = 0.7,
            bg.color = "white",
            bg.alpha = 0.7,
            position = tm_pos_in(
              "right", "bottom"),
            frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

### 1.9 Analysis by planning area

The tutorial compares Punggol, Tampines, Choa Chu Kang and Jurong West.

```{r}
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")
```

```{r}
#| fig-width: 16
#| fig-height: 12
par(mfrow=c(2,2))
plot(st_geometry(pg), main = "Ponggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
```

Each planning area is converted to its own observation window.

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

The Singapore point pattern is clipped to each window.

```{r}
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

```{r fig.width=16, fig.height=12}
par(mfrow=c(2,2))
plot(unmark(childcare_pg_ppp.km), 
  main="Punggol")
plot(unmark(childcare_tm_ppp.km), 
  main="Tampines")
plot(unmark(childcare_ck_ppp.km), 
  main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), 
  main="Jurong West")
```

The Clark--Evans tests below compare Choa Chu Kang and Tampines.

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

Finally, a separate automatic bandwidth is used to show the density pattern within each planning area.

```{r}
#| fig-width: 16
#| fig-height: 15
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Jurong West")
```

## Part 2: Second-order spatial point pattern analysis

Second-order analysis asks whether nearby childcare services are related at different distance ranges. The tutorial uses the G, F, K and L functions for Choa Chu Kang and Tampines.

### 2.1 Packages and data preparation

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

```{r}
mpsz_cl <- read_rds("chap04/data/mpsz_cl.rds")
```

```{r}
childcare_sf <- read_rds("chap04/data/childcare_sf.rds")
```

The tutorial uses `rjitter()` to move exactly coincident points by a very small random amount before the second-order analysis.

```{r}
childcare_ppp <- as.ppp(childcare_sf) %>%
  rjitter(retry = TRUE,
          nsim = 1,
          drop = TRUE)

sg_owin <- as.owin(mpsz_cl)
```

```{r}
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")

pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)

childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]

childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

```{r fig.width=16, fig.height=12}
par(mfrow=c(2,2))
plot(unmark(childcare_pg_ppp.km), main="Punggol")
plot(unmark(childcare_tm_ppp.km), main="Tampines")
plot(unmark(childcare_ck_ppp.km), main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), main="Jurong West")
```

The random seed makes the Monte Carlo results reproducible.

```{r}
set.seed(1234)
```

### 2.2 G-function: nearest-event distances

The G-function measures the distance from each childcare service to its nearest neighbouring service.

#### Choa Chu Kang

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

The envelope shows the range expected from 999 simulated random patterns.

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
plot(G_CK.csr)
```

#### Tampines

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
plot(G_tm.csr)
```

An observed curve above the random envelope at short distances means nearest neighbours occur sooner than expected, which suggests clustering.

### 2.3 F-function: empty-space distances

The F-function measures the distance from empty locations in the study area to the nearest childcare service.

#### Choa Chu Kang

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
plot(F_CK.csr)
```

#### Tampines

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
plot(F_tm.csr)
```

The G- and F-functions complement each other: G begins at service locations, while F begins in the empty space between them.

### 2.4 K-function: neighbours within distance $r$

The K-function counts neighbouring events within increasing distances. The tutorial plots K minus the distance line used as its CSR reference.

#### Choa Chu Kang

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)

```

```{r fig.width=8, fig.height=5}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

#### Tampines

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)

```

```{r fig.width=7,  fig.height=4.5}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

### 2.5 L-function: variance-stabilised K

The L-function is a transformed version of K that is easier to compare across distances.

#### Choa Chu Kang

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)

```

```{r fig.width=8, fig.height=5}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

#### Tampines

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)

```

```{r fig.width=7,  fig.height=5.5}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```

Together, the G, F, K and L plots show whether departures from randomness happen only among nearest neighbours or persist across a wider distance range.

## Some other questions I would like to address

After completing the tutorial, I would also like to ask:

1. Does an area with more services also have a higher concentration?
2. Do services sharing the same map location affect the nearest-neighbour result?
3. Where is the strongest concentration when every area uses the same smoothing distance?

### 3.1 Does having more services mean a higher concentration?

The four planning areas differ in size, so comparing totals alone is not enough. Here I compare the number of services per square kilometre. I use the original point coordinates, before the tutorial's random jitter, so shared locations can be checked in the next question.

```{r}
raw_childcare_ppp <- as.ppp(childcare_sf)

area_patterns <- list(
  "Punggol" = raw_childcare_ppp[pg_owin],
  "Tampines" = raw_childcare_ppp[tm_owin],
  "Choa Chu Kang" = raw_childcare_ppp[ck_owin],
  "Jurong West" = raw_childcare_ppp[jw_owin]
)

unique_location_pattern <- function(point_pattern) {
  unique_coordinate <- !duplicated(
    cbind(point_pattern$x, point_pattern$y)
  )
  point_pattern[unique_coordinate]
}

area_comparison <- imap_dfr(area_patterns, function(point_pattern, area_name) {
  unique_pattern <- unique_location_pattern(point_pattern)

  original_test <- clarkevans.test(
    point_pattern,
    correction = "none",
    alternative = "two.sided"
  )

  unique_test <- clarkevans.test(
    unique_pattern,
    correction = "none",
    alternative = "two.sided"
  )

  area_km2 <- area.owin(Window(point_pattern)) / 1e6

  tibble(
    planning_area = area_name,
    area_km2 = area_km2,
    services = npoints(point_pattern),
    unique_locations = npoints(unique_pattern),
    services_per_km2 = npoints(point_pattern) / area_km2,
    original_R = unname(original_test$statistic),
    unique_location_R = unname(unique_test$statistic),
    unique_location_p = unique_test$p.value
  )
})

area_comparison %>%
  transmute(
    `Planning area` = planning_area,
    `Area (km²)` = round(area_km2, 2),
    Services = services,
    `Services per km²` = round(services_per_km2, 2)
  ) %>%
  knitr::kable()
```

Tampines has the largest total, but Choa Chu Kang has the most services relative to its area. This is why a count and a density answer different questions.

### 3.2 Do shared map locations change the result?

When two records have exactly the same coordinates, their nearest-neighbour distance is zero. I repeat the Clark--Evans test after keeping one point at each location.

```{r}
area_comparison %>%
  transmute(
    `Planning area` = planning_area,
    `R: all records` = round(original_R, 3),
    `R: one point per location` = round(unique_location_R, 3),
    `Simple reading` = case_when(
      unique_location_p < 0.05 & unique_location_R < 1 ~ "Still appears clustered",
      TRUE ~ "No clear clustering"
    )
  ) %>%
  knitr::kable()
```

R near 1 looks more like randomness; lower values suggest stronger grouping. The chart shows how far each result moves after shared locations are counted once.

```{r}
#| fig-height: 5.5
ggplot(
  area_comparison,
  aes(y = reorder(planning_area, original_R))
) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    colour = "grey45"
  ) +
  geom_segment(
    aes(x = original_R, xend = unique_location_R, yend = planning_area),
    linewidth = 0.8,
    colour = "grey65"
  ) +
  geom_point(
    aes(x = original_R, colour = "All records", shape = "All records"),
    size = 3.2
  ) +
  geom_point(
    aes(
      x = unique_location_R,
      colour = "One point per location",
      shape = "One point per location"
    ),
    size = 3.2
  ) +
  scale_colour_manual(
    values = c(
      "All records" = "#d85b3f",
      "One point per location" = "#39735f"
    )
  ) +
  scale_shape_manual(
    values = c(
      "All records" = 16,
      "One point per location" = 17
    )
  ) +
  coord_cartesian(xlim = c(0.6, 1.04)) +
  labs(
    title = "Do shared map locations change the result?",
    subtitle = "R near 1 looks random; lower values suggest more grouping",
    x = "R value",
    y = NULL,
    colour = NULL,
    shape = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank()
  )
```

The comparison shows whether the clustering result is robust or mainly driven by several services recorded at the same address.

### 3.3 Where are the strongest concentrations?

The tutorial's planning-area maps select different bandwidths. For a direct comparison, I apply the same 0.6 km bandwidth to all four areas and locate each surface's peak.

```{r}
find_fixed_hotspot <- function(point_pattern, area_name) {
  point_pattern_km <- rescale.ppp(point_pattern, 1000, "km")

  density_surface <- density(
    point_pattern_km,
    sigma = 0.6,
    edge = TRUE,
    kernel = "gaussian"
  )

  peak_cell <- which(
    density_surface$v == max(density_surface$v, na.rm = TRUE),
    arr.ind = TRUE
  )[1, ]

  tibble(
    planning_area = area_name,
    peak_density = max(density_surface$v, na.rm = TRUE),
    easting = density_surface$xcol[peak_cell[2]] * 1000,
    northing = density_surface$yrow[peak_cell[1]] * 1000
  )
}

hotspot_points <- imap_dfr(
  area_patterns,
  ~ find_fixed_hotspot(.x, .y)
) %>%
  st_as_sf(coords = c("easting", "northing"), crs = 3414)

hotspot_comparison <- st_join(
  hotspot_points,
  mpsz_cl %>% select(SUBZONE_N),
  join = st_within
) %>%
  st_drop_geometry() %>%
  transmute(
    `Planning area` = planning_area,
    `Peak subzone` = str_to_title(SUBZONE_N),
    `Estimated services per km² at peak` = round(peak_density, 1)
  ) %>%
  arrange(desc(`Estimated services per km² at peak`))

hotspot_comparison %>%
  knitr::kable()
```

This comparison identifies where services are concentrated, but it cannot say whether provision is sufficient. That would also require enrolment capacity and local demand data.

### 3.4 Main takeaway

Counts, density surfaces and point-pattern tests describe different parts of the same story. The most useful conclusion is therefore not simply which area has more services, but whether its apparent clustering remains after area size and repeated locations are considered.

## References

- [Chapter 4: First-order spatial point pattern analysis](https://r4gdsa.netlify.app/chap04.html)
- [Chapter 5: Second-order spatial point pattern analysis](https://r4gdsa.netlify.app/chap05.html)
- [Child Care Services, ECDA](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view)
- [Master Plan 2019 Subzone Boundary (No Sea), URA](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view)

[← Back to Hands-on exercise](hands-on-exercise.qmd){.back-link}
