---
title: "Hands-on Exercise 1"
subtitle: "Geospatial data wrangling and choropleth mapping"
author: "Le Truong Chinh"
date: "3 September 2026"
description: "An executed ISSS626 exercise using sf, tidyverse and tmap with Singapore geospatial data."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 3
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 5.5
---

::: {.exercise-lead}
This exercise follows Chapters 1 and 2 of *R for Geospatial Data Science and Analytics*. Part 1 prepares and analyses spatial data. Part 2 turns population data into choropleth maps.
:::

## Part 1: Geospatial data wrangling

### 1.1 Data acquisition

The exercise uses four datasets. The original Master Plan 2014 “Web” file is no longer available from its collection, so the current official “No Sea” GeoJSON is used instead. This changes the import format, but not the analysis.

| Dataset | Format used | Source |
|---|---|---|
| Master Plan 2014 Subzone Boundary (No Sea) | GeoJSON polygon | [URA via data.gov.sg](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view) |
| Pre-Schools Location | GeoJSON point | [ECDA via data.gov.sg](https://data.gov.sg/datasets/d_61eefab99958fd70e6aab17320a71f1c/view) |
| Cycling Path Network | GeoJSON line | [LTA via data.gov.sg](https://data.gov.sg/datasets/d_8f468b25193f64be8a16fa7d8f60f553/view) |
| Singapore Airbnb listings, 29 June 2026 | CSV table | [Inside Airbnb](https://insideairbnb.com/get-the-data/) |

The files are stored in separate geospatial and aspatial folders.

```text
data/
├── geospatial/
│   ├── MasterPlan2014SubzoneBoundaryNoSea.geojson
│   ├── MasterPlan2019SubzoneBoundaryNoSea.geojson
│   ├── PreSchoolsLocation.geojson
│   └── CyclingPathNetwork.geojson
└── aspatial/
    ├── listings.csv
    └── respopagesextod2024.csv
```

### 1.2 Getting started

`sf` handles vector data, while packages in `tidyverse` import, transform and visualise the data. `tmap` is used in Part 2.

```{r}
#| label: setup
pacman::p_load(sf, tidyverse, tmap, classInt, knitr)

theme_set(theme_minimal(base_size = 12))
```

### 1.3 Importing geospatial data

#### Importing the polygon layer

`st_read()` imports the Master Plan 2014 subzone boundaries as a simple feature data frame named `mpsz`.

```{r}
#| label: import-mp14
mpsz <- st_read(
  "data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson"
)
```

The layer contains `r format(nrow(mpsz), big.mark = ",")` subzones and uses `r paste(unique(st_geometry_type(mpsz)), collapse = ", ")` geometry.

#### Importing the line layer

The cycling path network is imported in the same way.

```{r}
#| label: import-cycling
cyclingpath <- st_read(
  "data/geospatial/CyclingPathNetwork.geojson"
)
```

The current layer contains `r format(nrow(cyclingpath), big.mark = ",")` line features.

#### Importing the point layer

The pre-school locations form the point layer.

```{r}
#| label: import-preschools
preschool <- st_read(
  "data/geospatial/PreSchoolsLocation.geojson"
)
```

The file contains `r format(nrow(preschool), big.mark = ",")` pre-school points.

### 1.4 Checking a simple feature data frame

#### Working with `st_geometry()`

`st_geometry()` reports the geometry type, bounding box and coordinate reference system.

```{r}
#| label: inspect-geometry
st_geometry(mpsz)
```

#### Working with `glimpse()`

`glimpse()` shows the attribute names and data types without dropping the geometry.

```{r}
#| label: inspect-glimpse
glimpse(mpsz)
```

#### Working with `head()`

The first five records confirm that subzone, planning-area and region names are available for later analysis.

```{r}
#| label: inspect-head
head(mpsz, n = 5) |>
  select(SUBZONE_N, PLN_AREA_N, REGION_N, geometry)
```

### 1.5 Plotting geospatial data

A quick geometry plot checks the shape and extent of the subzone layer.

```{r}
#| label: plot-mp14-geometry
plot(
  st_geometry(mpsz),
  col = "#e7ece9",
  border = "white",
  main = "Master Plan 2014 subzones"
)
```

An attribute plot colours the polygons by planning area.

```{r}
#| label: plot-mp14-attribute
#| fig-height: 7
plot(
  mpsz["PLN_AREA_N"],
  key.pos = 4,
  main = "Singapore planning areas"
)
```

### 1.6 Working with projection

The current GeoJSON files are correctly labelled WGS 84 (EPSG:4326). They must be transformed to Singapore SVY21 (EPSG:3414) before calculating distances or areas.

```{r}
#| label: inspect-crs
tibble(
  layer = c("Subzones", "Cycling paths", "Pre-schools"),
  crs = c(
    st_crs(mpsz)$Name,
    st_crs(cyclingpath)$Name,
    st_crs(preschool)$Name
  )
) |>
  kable()
```

`st_transform()` recalculates the coordinates in the target CRS. `st_set_crs()` is not used because the source CRS is already correct; relabelling the coordinates would put the features in the wrong place.

```{r}
#| label: transform-crs
mpsz <- st_transform(mpsz, 3414)
cyclingpath <- st_transform(cyclingpath, 3414)
preschool <- st_transform(preschool, 3414)

tibble(
  layer = c("Subzones", "Cycling paths", "Pre-schools"),
  epsg = c(
    st_crs(mpsz)$epsg,
    st_crs(cyclingpath)$epsg,
    st_crs(preschool)$epsg
  )
) |>
  kable()
```

The transformed layers now share the same projected coordinate system and can be overlaid.

```{r}
#| label: overlay-preschools
ggplot() +
  geom_sf(data = mpsz, fill = "#edf1ef", colour = "white", linewidth = 0.15) +
  geom_sf(data = preschool, colour = "#d85b3f", size = 0.25, alpha = 0.65) +
  coord_sf(datum = NA) +
  labs(
    title = "Pre-school locations",
    subtitle = "Master Plan 2014 subzones"
  ) +
  theme_void()
```

### 1.7 Importing and converting aspatial data

#### Importing the Airbnb table

`read_csv()` imports the latest Singapore listings as a tibble.

```{r}
#| label: import-airbnb
listings <- read_csv(
  "data/aspatial/listings.csv",
  show_col_types = FALSE
)

glimpse(listings)
```

#### Creating a simple feature data frame

Longitude and latitude are WGS 84 coordinates. `st_as_sf()` creates point geometry, and `st_transform()` converts the result to SVY21.

```{r}
#| label: convert-airbnb
listings_sf <- listings |>
  drop_na(longitude, latitude) |>
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326
  ) |>
  st_transform(3414)

listings_sf
```

The converted object contains `r format(nrow(listings_sf), big.mark = ",")` listing points.

```{r}
#| label: map-airbnb
ggplot() +
  geom_sf(data = mpsz, fill = "#edf1ef", colour = "white", linewidth = 0.15) +
  geom_sf(data = listings_sf, colour = "#d85b3f", size = 0.35, alpha = 0.5) +
  coord_sf(datum = NA) +
  labs(
    title = "Airbnb listings in Singapore",
    subtitle = "Inside Airbnb snapshot, 29 June 2026"
  ) +
  theme_void()
```

The listings are concentrated in the central and south-eastern parts of Singapore rather than being evenly distributed across the island.

### 1.8 Geoprocessing with `sf`

#### Use case 1: land around cycling paths

A five-metre buffer marks the land within five metres of each cycling path segment.

```{r}
#| label: cycling-buffer
buffer_cycling <- st_buffer(
  cyclingpath,
  dist = 5,
  nQuadSegs = 30
)
```

```{r}
#| label: cycling-buffer-map
ggplot() +
  geom_sf(data = mpsz, fill = "#f4f5f2", colour = "white", linewidth = 0.1) +
  geom_sf(data = buffer_cycling, fill = "#d85b3f", colour = NA, alpha = 0.75) +
  coord_sf(datum = NA) +
  labs(title = "Five-metre cycling path buffer") +
  theme_void()
```

Summing the individual polygon areas gives the combined segment-level estimate.

```{r}
#| label: cycling-buffer-area
buffer_area_ha <- sum(st_area(buffer_cycling)) |>
  units::set_units(ha)

tibble(
  measure = "Sum of individual 5 m buffers",
  area_hectares = round(as.numeric(buffer_area_ha), 2)
) |>
  kable()
```

The estimate is `r round(as.numeric(buffer_area_ha), 2)` hectares. Overlapping buffers are counted more than once, so this is a segment-level total rather than a dissolved land footprint.

#### Use case 2: number of pre-schools by subzone

`st_intersects()` identifies which pre-school points fall inside each subzone. `lengths()` converts the list of matches into a count.

```{r}
#| label: preschool-count
mpsz$PreSch_count <- lengths(
  st_intersects(mpsz, preschool)
)

mpsz |>
  st_drop_geometry() |>
  select(SUBZONE_N, PLN_AREA_N, PreSch_count) |>
  arrange(desc(PreSch_count)) |>
  slice_head(n = 10) |>
  kable(
    col.names = c("Subzone", "Planning area", "Pre-schools"),
    caption = "Ten subzones with the most pre-schools"
  )
```

`r mpsz$SUBZONE_N[which.max(mpsz$PreSch_count)]` has the highest count, with `r max(mpsz$PreSch_count)` pre-schools.

### 1.9 Exploratory data analysis

#### Deriving area and density

Area is converted from square metres to square kilometres. Dividing the facility count by area gives pre-school density.

```{r}
#| label: preschool-density
mpsz <- mpsz |>
  mutate(
    Area_km2 = as.numeric(units::set_units(st_area(geometry), km^2)),
    PreSch_density = PreSch_count / Area_km2
  )

mpsz |>
  st_drop_geometry() |>
  select(SUBZONE_N, PLN_AREA_N, PreSch_count, Area_km2, PreSch_density) |>
  arrange(desc(PreSch_density)) |>
  slice_head(n = 10) |>
  mutate(
    Area_km2 = round(Area_km2, 2),
    PreSch_density = round(PreSch_density, 2)
  ) |>
  kable(
    col.names = c("Subzone", "Planning area", "Count", "Area (km²)", "Density per km²"),
    caption = "Ten subzones with the highest pre-school density"
  )
```

#### Histogram

```{r}
#| label: preschool-histogram
ggplot(mpsz, aes(x = PreSch_count)) +
  geom_histogram(binwidth = 5, boundary = 0, colour = "white", fill = "#39735f") +
  labs(
    title = "Distribution of pre-schools by subzone",
    x = "Number of pre-schools",
    y = "Number of subzones"
  )
```

The distribution is right-skewed: the median subzone contains `r median(mpsz$PreSch_count)` pre-schools, while the maximum is `r max(mpsz$PreSch_count)`.

#### Scatterplot

```{r}
#| label: preschool-scatterplot
ggplot(mpsz, aes(x = Area_km2, y = PreSch_count)) +
  geom_point(colour = "#d85b3f", alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE, colour = "#17333d") +
  labs(
    title = "Subzone area and pre-school count",
    x = "Subzone area (km²)",
    y = "Number of pre-schools"
  )
```

The fitted line provides a broad comparison, but the spread of points shows that subzone size alone does not explain the number of pre-schools.

## Part 2: Choropleth mapping

### 2.1 Getting started

Part 2 uses `tmap` for thematic mapping. Static plot mode keeps the maps suitable for the report page.

```{r}
#| label: tmap-setup
tmap_mode("plot")
```

### 2.2 Importing data

#### Importing Master Plan 2019 subzones

The original chapter uses KML and extracts attributes from an HTML description. That KML download is no longer available, so the current official GeoJSON is used. Its subzone attributes are already stored in separate columns.

```{r}
#| label: import-mp19
mpsz19 <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson"
) |>
  st_transform(3414)

mpsz19 |>
  select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C, geometry) |>
  head(10)
```

#### Importing resident population data

The 2024 SingStat data is stored as a CSV with planning area, subzone, age group, sex, population and year fields.

```{r}
#| label: import-population
popdata <- read_csv(
  "data/aspatial/respopagesextod2024.csv",
  na = c("na", "-", ""),
  show_col_types = FALSE
)

glimpse(popdata)
```

### 2.3 Preparing the population table

The dependency ratio compares young and aged residents with the economically active age group.

$$
\text{Dependency ratio} =
\frac{\text{Young} + \text{Aged}}
{\text{Economically active}}
$$

- **Young:** ages 0–24
- **Economically active:** ages 25–64
- **Aged:** ages 65 and above

The data is aggregated across sex and reshaped from age-group rows to columns.

```{r}
#| label: prepare-population
popdata2024 <- popdata |>
  filter(PA != "Total", SZ != "Total", AG != "Total") |>
  group_by(PA, SZ, AG) |>
  summarise(POP = sum(Pop, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = AG, values_from = POP, values_fill = 0) |>
  mutate(
    YOUNG = rowSums(across(c(
      `0 - 4`, `5 - 9`, `10 - 14`, `15 - 19`, `20 - 24`
    ))),
    `ECONOMY ACTIVE` = rowSums(across(c(
      `25 - 29`, `30 - 34`, `35 - 39`, `40 - 44`,
      `45 - 49`, `50 - 54`, `55 - 59`, `60 - 64`
    ))),
    AGED = rowSums(across(c(
      `65 - 69`, `70 - 74`, `75 - 79`, `80 - 84`,
      `85 - 89`, `90 & Over`
    ))),
    TOTAL = YOUNG + `ECONOMY ACTIVE` + AGED,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) |>
  mutate(across(c(PA, SZ), str_to_upper)) |>
  filter(`ECONOMY ACTIVE` > 0) |>
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)

popdata2024 |>
  arrange(desc(DEPENDENCY)) |>
  slice_head(n = 10) |>
  mutate(DEPENDENCY = round(DEPENDENCY, 2)) |>
  kable(
    col.names = c(
      "Planning area", "Subzone", "Young", "Economically active",
      "Aged", "Total", "Dependency ratio"
    ),
    caption = "Highest calculated dependency ratios"
  )
```

### 2.4 Joining attributes to subzones

The spatial layer stays on the left of `left_join()`, so the result remains an `sf` object. Upper-case subzone names provide the common key.

```{r}
#| label: join-population
mpsz_pop2024 <- mpsz19 |>
  mutate(SUBZONE_N = str_to_upper(SUBZONE_N)) |>
  left_join(
    popdata2024,
    by = c("SUBZONE_N" = "SZ")
  )

join_check <- mpsz_pop2024 |>
  st_drop_geometry() |>
  summarise(
    subzones = n(),
    matched = sum(!is.na(DEPENDENCY)),
    unmatched = sum(is.na(DEPENDENCY))
  )

join_check |>
  kable(caption = "Population-to-boundary join check")
```

The join matches `r join_check$matched` of `r join_check$subzones` subzones. The remaining `r join_check$unmatched` polygons have no dependency value, mainly because they have no resident population record or use a different subzone label.

```{r}
#| label: save-joined-data
dir.create("data/rds", showWarnings = FALSE)
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

### 2.5 Quick choropleth map

`qtm()` provides a fast first view of the dependency ratio.

```{r}
#| label: quick-map
qtm(
  shp = mpsz_pop2024,
  fill = "DEPENDENCY",
  title = "Dependency ratio by subzone"
)
```

### 2.6 Building a choropleth with `tmap`

`tm_shape()` defines the spatial object. `tm_polygons()` maps the dependency ratio to polygon fill, while `tm_borders()` separates neighbouring subzones.

```{r}
#| label: base-choropleth
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.legend = tm_legend(title = "Dependency ratio")
  ) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Dependency ratio by planning subzone")
```

### 2.7 Comparing classification methods

The class method changes which subzones appear similar. Equal intervals use equally sized numeric ranges; quantiles place similar numbers of subzones in each class; Jenks natural breaks follow clusters in the data.

```{r}
#| label: classification-comparison
#| fig-width: 12
#| fig-height: 4.8
map_equal <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "equal", n = 5, values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Equal")
  ) +
  tm_borders(col = "white", lwd = 0.3)

map_quantile <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 5, values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Quantile")
  ) +
  tm_borders(col = "white", lwd = 0.3)

map_jenks <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "jenks", n = 5, values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Jenks")
  ) +
  tm_borders(col = "white", lwd = 0.3)

tmap_arrange(map_equal, map_quantile, map_jenks, ncol = 3)
```

The equal-interval map is dominated by the low classes because a few large ratios stretch the range. Quantiles provide the clearest island-wide comparison, while Jenks highlights breaks in the observed distribution.

#### Comparing the number of classes

The maps below hold the quantile method constant and vary the number of classes.

```{r}
#| label: class-count-comparison
#| fig-width: 11
#| fig-height: 8
quantile_map <- function(classes) {
  tm_shape(mpsz_pop2024) +
    tm_polygons(
      fill = "DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = classes,
        values = "brewer.greens"
      ),
      fill.legend = tm_legend(title = paste(classes, "classes"))
    ) +
    tm_borders(col = "white", lwd = 0.25)
}

tmap_arrange(
  quantile_map(2),
  quantile_map(6),
  quantile_map(10),
  quantile_map(20),
  ncol = 2
)
```

Two classes remove most local variation. Twenty classes add detail but make the legend and spatial pattern difficult to read. Five or six classes provide a clearer balance.

### 2.8 Custom breaks and colour

The summary below provides the basis for manual class breaks.

```{r}
#| label: dependency-summary
summary(mpsz_pop2024$DEPENDENCY)
```

The custom map uses breaks at 0.60, 0.70, 0.80 and 0.90, with the final class extending to the observed maximum.

```{r}
#| label: custom-break-map
dependency_max <- max(mpsz_pop2024$DEPENDENCY, na.rm = TRUE)

tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.60, 0.70, 0.80, 0.90, dependency_max),
      values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Dependency ratio")
  ) +
  tm_borders(col = "white", lwd = 0.3)
```

A sequential green palette suits an ordered ratio: darker shades indicate higher values. Reversing the palette would make low values appear visually dominant, so the original direction is retained.

### 2.9 Final cartographic map

The final map uses six quantile classes, a concise title, border lines, a compass, a scale bar and a source note.

```{r}
#| label: final-map
#| fig-height: 6.5
final_map <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 6,
      values = "brewer.greens"
    ),
    fill.legend = tm_legend(
      title = "Dependency ratio",
      position = tm_pos_in("left", "bottom")
    )
  ) +
  tm_borders(col = "white", lwd = 0.35) +
  tm_title("Dependency ratio by Singapore planning subzone, 2024") +
  tm_compass(type = "8star", position = tm_pos_in("right", "top")) +
  tm_scalebar(position = tm_pos_in("right", "bottom")) +
  tm_credits(
    "Sources: URA Master Plan 2019; Singapore Department of Statistics",
    position = tm_pos_out("left", "bottom")
  ) +
  tm_layout(frame = FALSE)

final_map
```

The map shows the ratio, not the absolute number of dependants. A high value indicates more young and older residents relative to ages 25–64; it does not by itself measure service demand or household burden.

### 2.10 Small multiple maps

Young and aged population totals are mapped separately with the same classification method.

```{r}
#| label: small-multiples
#| fig-width: 11
#| fig-height: 5
young_map <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "YOUNG",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 5, values = "brewer.blues"
    ),
    fill.legend = tm_legend(title = "Young residents")
  ) +
  tm_borders(col = "white", lwd = 0.3)

aged_map <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "AGED",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 5, values = "brewer.oranges"
    ),
    fill.legend = tm_legend(title = "Aged residents")
  ) +
  tm_borders(col = "white", lwd = 0.3)

tmap_arrange(young_map, aged_map, ncol = 2)
```

The two maps separate population composition from the dependency ratio and make different concentrations visible.

### 2.11 Mapping a selected region

Filtering the `sf` object before mapping focuses the analysis on the Central Region.

```{r}
#| label: central-region-map
central_region <- mpsz_pop2024 |>
  filter(REGION_N == "CENTRAL REGION")

tm_shape(central_region) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 5, values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Dependency ratio")
  ) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Central Region dependency ratio")
```

### 2.12 Interactive map

Plot mode is switched to view mode for an interactive map. Hovering and clicking reveal the subzone name and dependency ratio.

```{r}
#| label: interactive-map
#| column: screen-inset-shaded
tmap_mode("view")

tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 6, values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Dependency ratio"),
    id = "SUBZONE_N",
    popup.vars = c(
      "Subzone" = "SUBZONE_N",
      "Planning area" = "PLN_AREA_N",
      "Dependency ratio" = "DEPENDENCY"
    )
  ) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_view(set_zoom_limits = c(10, 14))

tmap_mode("plot")
```

## Conclusion

The exercise shows that geospatial analysis depends on preparation before mapping. All layers must use a suitable CRS, spatial joins must be checked, and classification choices must be explained. The final map is therefore supported by the import checks, transformations and calculations shown above rather than presented as an isolated graphic.

[← Back to Hands-on exercise](hands-on-exercise.qmd){.back-link}
