---
title: "Hands-on Exercise 2"
subtitle: "First- and second-order spatial point pattern analysis"
author: "Le Truong Chinh"
date: "3 September 2026"
description: "An executed ISSS626 exercise analysing Singapore childcare centre locations with spatstat."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 3
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 5.5
---

::: {.exercise-lead}
This exercise follows Chapters 4 and 5 of *R for Geospatial Data Science and Analytics*. It examines the distribution of childcare centres in Singapore using first- and second-order spatial point pattern methods.
:::

## Part 1: First-order spatial point pattern analysis

### 1.1 Data and packages

Two official datasets are used:

| Dataset | Geometry | Source |
|---|---|---|
| Child Care Services | Point | [ECDA via data.gov.sg](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view) |
| Master Plan 2019 Subzone Boundary (No Sea) | Polygon | [URA via data.gov.sg](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view) |

The childcare file contains December 2021 records and was updated on 18 April 2026. GeoJSON is used instead of KML, so its attributes are already stored in separate fields and do not need to be extracted from an HTML `Description` field.

```{r}
#| label: setup
library(sf)
library(terra)
library(spatstat)
library(tmap)
library(tidyverse)

options(width = 90)
tmap_mode("plot")
```

### 1.2 Importing and preparing the geospatial data

Both layers are imported with `st_read()`, transformed to Singapore SVY21 / EPSG:3414 and checked for valid geometry. The two offshore planning areas excluded in the course tutorial are also removed.

```{r}
#| label: import-data
mpsz_sf <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson",
  quiet = TRUE
) |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(3414) |>
  st_make_valid()

mpsz_cl <- mpsz_sf |>
  filter(
    PLN_AREA_N != "WESTERN ISLANDS",
    PLN_AREA_N != "NORTH-EASTERN ISLANDS"
  )

childcare_sf <- st_read(
  "data/geospatial/ChildCareServices.geojson",
  quiet = TRUE
) |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(3414)
```

The imported layers have the expected geometry types and a common projected coordinate reference system.

```{r}
#| label: check-imported-data
tibble(
  layer = c("Child care services", "Planning subzones"),
  features = c(nrow(childcare_sf), nrow(mpsz_cl)),
  geometry = c(
    as.character(unique(st_geometry_type(childcare_sf))),
    paste(unique(st_geometry_type(mpsz_cl)), collapse = ", ")
  ),
  epsg = c(st_crs(childcare_sf)$epsg, st_crs(mpsz_cl)$epsg)
) |>
  knitr::kable()
```

There are `r nrow(childcare_sf)` childcare service records. `r sum(duplicated(st_coordinates(childcare_sf)))` records share coordinates with an earlier record, usually because multiple services operate at the same address. They are retained as separate service events, but this should be considered when interpreting nearest-neighbour results.

### 1.3 Mapping the two layers

A static map confirms that the childcare points and planning boundaries occupy the same spatial extent.

```{r}
#| label: map-imported-data
#| fig-cap: "Childcare service locations within Singapore planning subzones."
tm_shape(mpsz_cl) +
  tm_polygons(fill = "#edf1ed", col = "#80938c", lwd = 0.5) +
  tm_shape(childcare_sf) +
  tm_dots(fill = "#d85b3f", col = "white", size = 0.025, lwd = 0.2) +
  tm_layout(frame = FALSE)
```

The same points can be explored interactively. The map is returned to static plot mode immediately after it is created.

```{r}
#| label: interactive-childcare-map
#| fig-cap: "Interactive map of childcare service locations."
tmap_mode("view")

childcare_view <- tm_shape(mpsz_cl) +
  tm_borders(col = "#748883", lwd = 1) +
  tm_shape(childcare_sf) +
  tm_dots(
    fill = "#d85b3f",
    size = 0.03,
    popup.vars = c("Centre" = "NAME", "Address" = "ADDRESSSTREETNAME")
  )

childcare_view
tmap_mode("plot")
```

### 1.4 Converting `sf` objects to `ppp` and `owin`

`spatstat` represents point events with a planar point pattern (`ppp`) and the study boundary with an observation window (`owin`). The childcare attributes are removed because this is an unmarked point pattern analysis.

```{r}
#| label: create-ppp
childcare_ppp <- as.ppp(childcare_sf) |>
  unmark()

class(childcare_ppp)
summary(childcare_ppp)
```

The subzones are dissolved into a single Singapore study boundary before conversion to `owin`. This removes internal subzone borders from the observation window.

```{r}
#| label: create-owin
sg_boundary <- st_union(mpsz_cl)
sg_owin <- as.owin(sg_boundary)

class(sg_owin)
plot(sg_owin, main = "Singapore observation window")
```

Only childcare events inside the Singapore window are retained.

```{r}
#| label: combine-ppp-owin
childcareSG_ppp <- childcare_ppp[sg_owin]
childcareSG_ppp
```

### 1.5 Clark–Evans nearest-neighbour test

The Clark–Evans test evaluates whether nearest-neighbour distances are shorter than expected under complete spatial randomness (CSR).

- $H_0$: childcare services follow CSR.
- $H_1$: childcare services are more clustered than CSR.

First, the asymptotic Z-test is performed without edge correction.

```{r}
#| label: clark-evans-z
clark_sg_z <- clarkevans.test(
  childcareSG_ppp,
  correction = "none",
  alternative = "clustered"
)

clark_sg_z
```

Next, the p-value is estimated from 99 CSR simulations while holding the number of points and the observation window constant.

```{r}
#| label: clark-evans-monte-carlo
set.seed(626)

clark_sg_mc <- clarkevans.test(
  childcareSG_ppp,
  correction = "none",
  alternative = "clustered",
  method = "MonteCarlo",
  nsim = 99
)

clark_sg_mc
```

The Monte Carlo test gives $R = `r round(unname(clark_sg_mc$statistic), 3)`$ and $p = `r format.pval(clark_sg_mc$p.value, digits = 3)`$. At the 5% level, the CSR hypothesis is `r if (clark_sg_mc$p.value < 0.05) "rejected" else "not rejected"`.

### 1.6 Kernel density estimation

Kernel density estimation (KDE) transforms the childcare points into a continuous intensity surface. The first surface uses Diggle's automatic bandwidth, a Gaussian kernel and edge correction.

```{r}
#| label: kde-diggle-metres
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)

plot(kde_SG_diggle, main = "KDE using bw.diggle (points per m²)")
summary(kde_SG_diggle)
```

The selected bandwidth is shown in metres.

```{r}
#| label: diggle-bandwidth-metres
bw_m <- bw.diggle(childcareSG_ppp)
bw_m
```

Rescaling the point pattern to kilometres produces identical spatial shapes and more readable density values.

```{r}
#| label: rescale-to-kilometres
childcareSG_ppp_km <- rescale.ppp(childcareSG_ppp, 1000, "km")

kde_childcareSG_km <- density(
  childcareSG_ppp_km,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)

plot(kde_childcareSG_km, main = "KDE using bw.diggle (points per km²)")
```

#### Comparing automatic bandwidth selectors

Different selectors make different assumptions about the point pattern. Scott's selector returns separate bandwidths for the x and y directions.

```{r}
#| label: compare-bandwidths
bw_cvl <- bw.CvL(childcareSG_ppp_km)
bw_scott <- bw.scott(childcareSG_ppp_km)
bw_ppl_value <- bw.ppl(childcareSG_ppp_km)
bw_diggle_value <- bw.diggle(childcareSG_ppp_km)

tibble(
  method = c("bw.CvL", "bw.scott", "bw.ppl", "bw.diggle"),
  bandwidth_x_km = c(
    as.numeric(bw_cvl),
    as.numeric(bw_scott[1]),
    as.numeric(bw_ppl_value),
    as.numeric(bw_diggle_value)
  ),
  bandwidth_y_km = c(
    as.numeric(bw_cvl),
    as.numeric(bw_scott[2]),
    as.numeric(bw_ppl_value),
    as.numeric(bw_diggle_value)
  )
) |>
  mutate(across(starts_with("bandwidth"), ~ round(.x, 3))) |>
  knitr::kable()
```

The smaller Diggle bandwidth preserves tighter local peaks, while the larger likelihood cross-validation bandwidth gives a smoother surface.

```{r}
#| label: compare-diggle-ppl
#| fig-width: 12
#| fig-height: 5
kde_childcareSG_ppl <- density(
  childcareSG_ppp_km,
  sigma = bw.ppl,
  edge = TRUE,
  kernel = "gaussian"
)

par(mfrow = c(1, 2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG_ppl, main = "bw.ppl")
par(mfrow = c(1, 1))
```

#### Comparing kernel functions

The same Diggle bandwidth is used so that only the kernel shape changes.

```{r}
#| label: compare-kernels
#| fig-width: 12
#| fig-height: 10
par(mfrow = c(2, 2))

for (kernel_name in c("gaussian", "epanechnikov", "quartic", "disc")) {
  plot(
    density(
      childcareSG_ppp_km,
      sigma = as.numeric(bw_diggle_value),
      edge = TRUE,
      kernel = kernel_name
    ),
    main = tools::toTitleCase(kernel_name)
  )
}

par(mfrow = c(1, 1))
```

### 1.7 Fixed and adaptive KDE

A fixed bandwidth of 0.6 km applies the same smoothing scale everywhere. Adaptive KDE changes the local bandwidth in response to the distribution of observations.

```{r}
#| label: fixed-adaptive-kde
#| fig-width: 12
#| fig-height: 5
kde_childcareSG_fixed <- density(
  childcareSG_ppp_km,
  sigma = 0.6,
  edge = TRUE,
  kernel = "gaussian"
)

kde_childcareSG_adaptive <- adaptive.density(
  childcareSG_ppp_km,
  method = "kernel",
  dimyx = 96
)

par(mfrow = c(1, 2))
plot(kde_childcareSG_fixed, main = "Fixed bandwidth: 0.6 km")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth")
par(mfrow = c(1, 1))
```

### 1.8 A cartographic KDE map

The metre-based KDE image is converted to a `SpatRaster`. Its values are multiplied by one million to change points per m² into points per km², while its SVY21 coordinates remain in metres.

```{r}
#| label: kde-to-terra
kde_childcare_terra <- rast(kde_SG_diggle)
values(kde_childcare_terra) <- values(kde_childcare_terra) * 1e6
names(kde_childcare_terra) <- "centres_per_km2"
crs(kde_childcare_terra) <- "EPSG:3414"

kde_childcare_terra
```

```{r}
#| label: cartographic-kde
#| fig-height: 6.5
#| fig-cap: "Kernel density of childcare services in centres per square kilometre."
tm_shape(kde_childcare_terra) +
  tm_raster(
    col.scale = tm_scale_continuous(values = "viridis"),
    col.legend = tm_legend(title = "Centres per km²")
  ) +
  tm_shape(sg_boundary) +
  tm_borders(col = "white", lwd = 1) +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_layout(frame = FALSE)
```

### 1.9 Analysis by planning area

Following the tutorial, four planning areas are extracted and dissolved into separate study windows.

```{r}
#| label: extract-planning-areas
pg <- mpsz_cl |>
  filter(PLN_AREA_N == "PUNGGOL") |>
  summarise()

tm <- mpsz_cl |>
  filter(PLN_AREA_N == "TAMPINES") |>
  summarise()

ck <- mpsz_cl |>
  filter(PLN_AREA_N == "CHOA CHU KANG") |>
  summarise()

jw <- mpsz_cl |>
  filter(PLN_AREA_N == "JURONG WEST") |>
  summarise()

par(mfrow = c(2, 2))
plot(st_geometry(pg), main = "Punggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
par(mfrow = c(1, 1))
```

Each planning area becomes an `owin`, which is then used to clip the national point pattern.

```{r}
#| label: planning-area-patterns
pg_owin <- as.owin(pg)
tm_owin <- as.owin(tm)
ck_owin <- as.owin(ck)
jw_owin <- as.owin(jw)

childcare_pg_ppp <- childcare_ppp[pg_owin]
childcare_tm_ppp <- childcare_ppp[tm_owin]
childcare_ck_ppp <- childcare_ppp[ck_owin]
childcare_jw_ppp <- childcare_ppp[jw_owin]

childcare_pg_ppp_km <- rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp_km <- rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp_km <- rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp_km <- rescale.ppp(childcare_jw_ppp, 1000, "km")

tibble(
  planning_area = c("Punggol", "Tampines", "Choa Chu Kang", "Jurong West"),
  childcare_services = c(
    npoints(childcare_pg_ppp),
    npoints(childcare_tm_ppp),
    npoints(childcare_ck_ppp),
    npoints(childcare_jw_ppp)
  )
) |>
  knitr::kable()
```

```{r}
#| label: plot-planning-area-patterns
#| fig-width: 12
#| fig-height: 10
par(mfrow = c(2, 2))
plot(childcare_pg_ppp_km, main = "Punggol")
plot(childcare_tm_ppp_km, main = "Tampines")
plot(childcare_ck_ppp_km, main = "Choa Chu Kang")
plot(childcare_jw_ppp_km, main = "Jurong West")
par(mfrow = c(1, 1))
```

Clark–Evans tests are run for Choa Chu Kang and Tampines, the two areas used again in Part 2.

```{r}
#| label: planning-area-clark-evans
clark_ck <- clarkevans.test(
  childcare_ck_ppp,
  correction = "none",
  alternative = "two.sided"
)

clark_tm <- clarkevans.test(
  childcare_tm_ppp,
  correction = "none",
  alternative = "two.sided"
)

tibble(
  planning_area = c("Choa Chu Kang", "Tampines"),
  R = c(unname(clark_ck$statistic), unname(clark_tm$statistic)),
  p_value = c(clark_ck$p.value, clark_tm$p.value),
  decision_at_5_percent = if_else(
    p_value < 0.05,
    "Reject CSR",
    "Do not reject CSR"
  )
) |>
  mutate(across(c(R, p_value), ~ round(.x, 4))) |>
  knitr::kable()
```

Finally, Diggle-bandwidth KDE surfaces are computed separately for all four planning areas.

```{r}
#| label: planning-area-kde
#| fig-width: 12
#| fig-height: 10
par(mfrow = c(2, 2))

plot(
  density(childcare_pg_ppp_km, sigma = bw.diggle, edge = TRUE),
  main = "Punggol"
)
plot(
  density(childcare_tm_ppp_km, sigma = bw.diggle, edge = TRUE),
  main = "Tampines"
)
plot(
  density(childcare_ck_ppp_km, sigma = bw.diggle, edge = TRUE),
  main = "Choa Chu Kang"
)
plot(
  density(childcare_jw_ppp_km, sigma = bw.diggle, edge = TRUE),
  main = "Jurong West"
)

par(mfrow = c(1, 1))
```

## Part 2: Second-order spatial point pattern analysis

Second-order methods examine how points relate to one another across a range of distances. Choa Chu Kang and Tampines are compared using G, F, K and L functions.

For every test:

- $H_0$: childcare services follow CSR within the planning area.
- $H_1$: childcare services do not follow CSR within the planning area.

### 2.1 G-function: nearest-event distances

The G-function describes the distance from each observed event to its nearest neighbouring event. A curve above the theoretical CSR curve indicates more short nearest-neighbour distances.

```{r}
#| label: g-function
#| fig-width: 12
#| fig-height: 5
G_ck <- Gest(childcare_ck_ppp, correction = "border")
G_tm <- Gest(childcare_tm_ppp, correction = "best")

par(mfrow = c(1, 2))
plot(G_ck, main = "G-function: Choa Chu Kang")
plot(G_tm, main = "G-function: Tampines")
par(mfrow = c(1, 1))
```

Pointwise CSR envelopes are generated from 999 simulations, matching the simulation count in the tutorial.

```{r}
#| label: g-envelopes
#| fig-width: 12
#| fig-height: 5
set.seed(626)
G_ck_csr <- envelope(
  childcare_ck_ppp,
  Gest,
  correction = "border",
  nsim = 999,
  rank = 1,
  verbose = FALSE
)

set.seed(626)
G_tm_csr <- envelope(
  childcare_tm_ppp,
  Gest,
  correction = "best",
  nsim = 999,
  rank = 1,
  verbose = FALSE
)

par(mfrow = c(1, 2))
plot(G_ck_csr, main = "G CSR envelope: Choa Chu Kang")
plot(G_tm_csr, main = "G CSR envelope: Tampines")
par(mfrow = c(1, 1))
```

### 2.2 F-function: empty-space distances

The F-function measures the distance from arbitrary locations in the window to the nearest observed event. A curve below CSR is consistent with larger empty spaces between clusters.

```{r}
#| label: f-function
#| fig-width: 12
#| fig-height: 5
F_ck <- Fest(childcare_ck_ppp, correction = "best")
F_tm <- Fest(childcare_tm_ppp, correction = "best")

par(mfrow = c(1, 2))
plot(F_ck, main = "F-function: Choa Chu Kang")
plot(F_tm, main = "F-function: Tampines")
par(mfrow = c(1, 1))
```

```{r}
#| label: f-envelopes
#| fig-width: 12
#| fig-height: 5
set.seed(626)
F_ck_csr <- envelope(
  childcare_ck_ppp,
  Fest,
  correction = "best",
  nsim = 999,
  rank = 1,
  verbose = FALSE
)

set.seed(626)
F_tm_csr <- envelope(
  childcare_tm_ppp,
  Fest,
  correction = "best",
  nsim = 999,
  rank = 1,
  verbose = FALSE
)

par(mfrow = c(1, 2))
plot(F_ck_csr, main = "F CSR envelope: Choa Chu Kang")
plot(F_tm_csr, main = "F CSR envelope: Tampines")
par(mfrow = c(1, 1))
```

### 2.3 K-function: neighbours within distance $r$

Ripley's K-function counts neighbours within increasing distances. Plotting $K(r)-K_{CSR}(r)$ centres the theoretical CSR expectation on zero; positive values indicate more neighbours than expected.

```{r}
#| label: k-function
#| fig-width: 12
#| fig-height: 5
K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")
K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")

par(mfrow = c(1, 2))
plot(K_ck, . - theo ~ r, main = "K - K(CSR): Choa Chu Kang")
abline(h = 0, lty = 2, col = "grey40")
plot(K_tm, . - theo ~ r, main = "K - K(CSR): Tampines")
abline(h = 0, lty = 2, col = "grey40")
par(mfrow = c(1, 1))
```

The K envelopes use 99 simulations and a rank-one global envelope, as in the tutorial. `global = TRUE` is used in place of the tutorial's misspelled `glocal` argument.

```{r}
#| label: k-envelopes
#| fig-width: 12
#| fig-height: 5
set.seed(626)
K_ck_csr <- envelope(
  childcare_ck_ppp,
  Kest,
  correction = "Ripley",
  nsim = 99,
  rank = 1,
  global = TRUE,
  verbose = FALSE
)

set.seed(626)
K_tm_csr <- envelope(
  childcare_tm_ppp,
  Kest,
  correction = "Ripley",
  nsim = 99,
  rank = 1,
  global = TRUE,
  verbose = FALSE
)

par(mfrow = c(1, 2))
plot(K_ck_csr, . - theo ~ r, main = "K global envelope: Choa Chu Kang")
abline(h = 0, lty = 2, col = "grey40")
plot(K_tm_csr, . - theo ~ r, main = "K global envelope: Tampines")
abline(h = 0, lty = 2, col = "grey40")
par(mfrow = c(1, 1))
```

### 2.4 L-function: variance-stabilised K

The L-function transforms K to make departures from CSR easier to compare across distances. Under CSR, $L(r)-r=0$.

```{r}
#| label: l-function
#| fig-width: 12
#| fig-height: 5
L_ck <- Lest(childcare_ck_ppp, correction = "Ripley")
L_tm <- Lest(childcare_tm_ppp, correction = "Ripley")

par(mfrow = c(1, 2))
plot(L_ck, . - theo ~ r, main = "L - L(CSR): Choa Chu Kang")
abline(h = 0, lty = 2, col = "grey40")
plot(L_tm, . - theo ~ r, main = "L - L(CSR): Tampines")
abline(h = 0, lty = 2, col = "grey40")
par(mfrow = c(1, 1))
```

```{r}
#| label: l-envelopes
#| fig-width: 12
#| fig-height: 5
set.seed(626)
L_ck_csr <- envelope(
  childcare_ck_ppp,
  Lest,
  correction = "Ripley",
  nsim = 99,
  rank = 1,
  global = TRUE,
  verbose = FALSE
)

set.seed(626)
L_tm_csr <- envelope(
  childcare_tm_ppp,
  Lest,
  correction = "Ripley",
  nsim = 99,
  rank = 1,
  global = TRUE,
  verbose = FALSE
)

par(mfrow = c(1, 2))
plot(L_ck_csr, . - theo ~ r, main = "L global envelope: Choa Chu Kang")
abline(h = 0, lty = 2, col = "grey40")
plot(L_tm_csr, . - theo ~ r, main = "L global envelope: Tampines")
abline(h = 0, lty = 2, col = "grey40")
par(mfrow = c(1, 1))
```

### 2.5 CSR envelope summary

The table records whether each observed function leaves its simulated CSR envelope at any evaluated distance. Leaving an envelope is evidence against CSR at one or more spatial scales; the plots show the direction and distance range of the departure.

```{r}
#| label: envelope-summary
summarise_envelope <- function(envelope_object, function_name, area_name) {
  valid <- with(
    envelope_object,
    is.finite(obs) & is.finite(lo) & is.finite(hi) & is.finite(theo)
  )

  delta <- envelope_object$obs[valid] - envelope_object$theo[valid]
  distance <- envelope_object$r[valid]
  strongest <- which.max(abs(delta))

  tibble(
    statistic = function_name,
    planning_area = area_name,
    leaves_CSR_envelope = any(
      envelope_object$obs[valid] < envelope_object$lo[valid] |
        envelope_object$obs[valid] > envelope_object$hi[valid]
    ),
    strongest_departure = if_else(delta[strongest] > 0, "Above CSR", "Below CSR"),
    distance_metres = round(distance[strongest])
  )
}

bind_rows(
  summarise_envelope(G_ck_csr, "G", "Choa Chu Kang"),
  summarise_envelope(G_tm_csr, "G", "Tampines"),
  summarise_envelope(F_ck_csr, "F", "Choa Chu Kang"),
  summarise_envelope(F_tm_csr, "F", "Tampines"),
  summarise_envelope(K_ck_csr, "K", "Choa Chu Kang"),
  summarise_envelope(K_tm_csr, "K", "Tampines"),
  summarise_envelope(L_ck_csr, "L", "Choa Chu Kang"),
  summarise_envelope(L_tm_csr, "L", "Tampines")
) |>
  knitr::kable()
```

The four functions describe complementary aspects of the same point pattern. G focuses on event-to-event spacing, F on empty space, and K/L on cumulative interaction across distance. Results should therefore be interpreted together rather than from a single curve.

## Some other questions I would like to address

After looking at the tutorial results, I have three more questions:

1. Does having more childcare services mean an area has a higher concentration?
2. Do services that share the same map location change the result?
3. Where is the main concentration in each planning area when I use the same map setting?

### 3.1 Does having more services mean a higher concentration?

The four planning areas have different sizes, so I compare both the total number of services and the number per square kilometre. I also prepare a second version of the data with only one point at each map location. The second version is used in the next question.

```{r}
#| label: cross-area-sensitivity
area_patterns <- list(
  "Punggol" = childcare_pg_ppp,
  "Tampines" = childcare_tm_ppp,
  "Choa Chu Kang" = childcare_ck_ppp,
  "Jurong West" = childcare_jw_ppp
)

unique_location_pattern <- function(point_pattern) {
  unique_coordinate <- !duplicated(
    cbind(point_pattern$x, point_pattern$y)
  )
  point_pattern[unique_coordinate]
}

area_comparison <- imap_dfr(area_patterns, function(point_pattern, area_name) {
  unique_pattern <- unique_location_pattern(point_pattern)

  original_test <- clarkevans.test(
    point_pattern,
    correction = "none",
    alternative = "two.sided"
  )

  unique_test <- clarkevans.test(
    unique_pattern,
    correction = "none",
    alternative = "two.sided"
  )

  area_km2 <- area.owin(Window(point_pattern)) / 1e6

  tibble(
    planning_area = area_name,
    area_km2 = area_km2,
    services = npoints(point_pattern),
    unique_locations = npoints(unique_pattern),
    services_per_km2 = npoints(point_pattern) / area_km2,
    original_R = unname(original_test$statistic),
    unique_location_R = unname(unique_test$statistic),
    unique_location_p = unique_test$p.value
  )
})

area_comparison |>
  transmute(
    `Planning area` = planning_area,
    `Area (km²)` = round(area_km2, 2),
    Services = services,
    `Services per km²` = round(services_per_km2, 2)
  ) |>
  knitr::kable()
```

Tampines has the most services, with `r area_comparison |> filter(planning_area == "Tampines") |> pull(services)`. Choa Chu Kang has only `r area_comparison |> filter(planning_area == "Choa Chu Kang") |> pull(services)`, but it is much smaller. As a result, Choa Chu Kang has the highest concentration at `r round(area_comparison |> filter(planning_area == "Choa Chu Kang") |> pull(services_per_km2), 2)` services per km². This shows why the total count alone can be misleading.

### 3.2 Do shared map locations change the result?

Some services have exactly the same coordinates, possibly because they are located at the same address or building. The nearest-neighbour test treats the distance between these records as zero, which can make the points look strongly grouped.

I run the test again using one point per map location. The test produces a value called **R**:

- R close to 1 means the points look roughly random.
- R below 1 means the points are closer together than expected.

```{r}
#| label: shared-location-results
area_comparison |>
  transmute(
    `Planning area` = planning_area,
    `R: all records` = round(original_R, 3),
    `R: one point per location` = round(unique_location_R, 3),
    `Simple reading` = case_when(
      unique_location_p < 0.05 & unique_location_R < 1 ~ "Still appears clustered",
      TRUE ~ "No clear clustering"
    )
  ) |>
  knitr::kable()
```

The chart shows the same comparison visually. A point moving towards the dashed line means the pattern looks less clustered after shared locations are counted once.

```{r}
#| label: clark-evans-sensitivity-chart
#| fig-height: 5.5
#| fig-cap: "How the result changes when shared map locations are counted once."
ggplot(
  area_comparison,
  aes(y = reorder(planning_area, original_R))
) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    colour = "grey45"
  ) +
  geom_segment(
    aes(x = original_R, xend = unique_location_R, yend = planning_area),
    linewidth = 0.8,
    colour = "grey65"
  ) +
  geom_point(
    aes(x = original_R, colour = "All records", shape = "All records"),
    size = 3.2
  ) +
  geom_point(
    aes(
      x = unique_location_R,
      colour = "One point per location",
      shape = "One point per location"
    ),
    size = 3.2
  ) +
  scale_colour_manual(
    values = c(
      "All records" = "#d85b3f",
      "One point per location" = "#39735f"
    )
  ) +
  scale_shape_manual(
    values = c(
      "All records" = 16,
      "One point per location" = 17
    )
  ) +
  coord_cartesian(xlim = c(0.6, 1.04)) +
  labs(
    title = "Do shared map locations change the result?",
    subtitle = "R near 1 looks random; lower values suggest more grouping",
    x = "R value",
    y = NULL,
    colour = NULL,
    shape = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank()
  )
```

Choa Chu Kang changes the most. Its R value moves from `r round(area_comparison |> filter(planning_area == "Choa Chu Kang") |> pull(original_R), 3)` to `r round(area_comparison |> filter(planning_area == "Choa Chu Kang") |> pull(unique_location_R), 3)`. Once repeated locations are counted once, its points no longer show clear clustering. This suggests that much of the original result comes from several services sharing the same location.

Tampines still looks clustered after the same adjustment, with R = `r round(area_comparison |> filter(planning_area == "Tampines") |> pull(unique_location_R), 3)`. Jurong West also remains clustered, while Punggol no longer shows a clear pattern.

The G, F, K and L charts tell a similar story. In Tampines, all four observed curves move outside the simulated random ranges. Choa Chu Kang is less clear: its F curve stays inside the random range, while its strongest G and L differences occur at about one or two metres. Those very short distances are likely related to services sharing coordinates.

### 3.3 Where are the strongest concentrations?

The earlier KDE maps choose a different smoothing distance for each planning area. That is useful for viewing each area separately, but it makes the peak values difficult to compare. Here, I use the same 600-metre smoothing distance for all four areas.

```{r}
#| label: fixed-bandwidth-hotspots
find_fixed_hotspot <- function(point_pattern, area_name) {
  point_pattern_km <- rescale.ppp(point_pattern, 1000, "km")

  density_surface <- density(
    point_pattern_km,
    sigma = 0.6,
    edge = TRUE,
    kernel = "gaussian"
  )

  peak_cell <- which(
    density_surface$v == max(density_surface$v, na.rm = TRUE),
    arr.ind = TRUE
  )[1, ]

  tibble(
    planning_area = area_name,
    peak_density = max(density_surface$v, na.rm = TRUE),
    easting = density_surface$xcol[peak_cell[2]] * 1000,
    northing = density_surface$yrow[peak_cell[1]] * 1000
  )
}

hotspot_points <- imap_dfr(
  area_patterns,
  ~ find_fixed_hotspot(.x, .y)
) |>
  st_as_sf(coords = c("easting", "northing"), crs = 3414)

hotspot_comparison <- st_join(
  hotspot_points,
  mpsz_cl |> select(SUBZONE_N),
  join = st_within
) |>
  st_drop_geometry() |>
  transmute(
    `Planning area` = planning_area,
    `Peak subzone` = str_to_title(SUBZONE_N),
    `Estimated services per km² at peak` = round(peak_density, 1)
  ) |>
  arrange(desc(`Estimated services per km² at peak`))

hotspot_comparison |>
  knitr::kable()
```

Using the same map setting, the strongest concentration appears in Taman Jurong, followed by Peng Siang, Punggol Field and Tampines West. This only tells me where services are concentrated. It does not tell me whether an area has enough childcare places because the data does not include enrolment capacity or demand.

### 3.4 What do I take away from this?

The total number of services does not give the whole picture. Choa Chu Kang has the highest number of services relative to its size, but its apparent clustering mostly disappears when repeated map locations are counted once. Tampines has a lower number per km², yet its service locations remain clearly grouped even after that adjustment.

From these results, I would describe Tampines as having the clearer clustering pattern. For Choa Chu Kang, I would be more careful: the result is strongly affected by services that share the same mapped location.

## References

- [Chapter 4: First-order spatial point pattern analysis](https://r4gdsa.netlify.app/chap04.html)
- [Chapter 5: Second-order spatial point pattern analysis](https://r4gdsa.netlify.app/chap05.html)
- [Child Care Services, ECDA](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view)
- [Master Plan 2019 Subzone Boundary (No Sea), URA](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view)

[← Back to Hands-on exercise](hands-on-exercise.qmd){.back-link}
