---
title: "Hands-on Exercise 1"
subtitle: "Geospatial data wrangling and choropleth mapping"
author: "Le Truong Chinh"
date: "3 September 2026"
description: "An executed ISSS626 exercise using sf, tidyverse and tmap with Singapore geospatial data."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 3
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 5.5
---

::: {.exercise-lead}
This exercise follows Chapters 1 and 2 of *R for Geospatial Data Science and Analytics*. Each tutorial code cell is kept separate and shown with its result.
:::

## Part 1: Geospatial data wrangling

This section follows Chapter 1. It covers data import, coordinate systems, conversion, geoprocessing and simple exploratory analysis.

### 1.1 Packages

```{r}
pacman::p_load(sf, tidyverse)
```

### 1.2 Importing geospatial data

The tutorial first imports the Master Plan 2014 subzones from an ESRI Shapefile.

```{r}
mpsz = st_read(dsn = "chap01/data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

The cycling path layer is imported in the same way.

```{r}
cyclingpath = st_read(dsn = "chap01/data/geospatial", 
                         layer = "CyclingPathGazette")
```

The preschool layer is stored as KML. It is transformed to Singapore SVY21 after import.

```{r}
preschool = st_read("chap01/data/geospatial/PreSchoolsLocation.kml") %>% 
  st_transform(crs = 3414)
```

### 1.3 Checking a simple feature data frame

`st_geometry()` displays the geometry column and its spatial properties.

```{r}
st_geometry(mpsz)
```

`glimpse()` gives a compact view of the fields and their data types.

```{r}
glimpse(mpsz)
```

```{r}
head(mpsz, n=5)  
```

### 1.4 Plotting geospatial data

Plotting the complete `sf` object creates a small map for each attribute.

```{r}
plot(mpsz)
```

Only the geometry can be plotted when the attributes are not needed.

```{r}
plot(st_geometry(mpsz))
```

One selected attribute can also be mapped.

```{r}
plot(mpsz["PLN_AREA_N"])
```

The two layers can be placed on the same map because they now use the same coordinate system.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), 
     add = TRUE)
```

### 1.5 Working with projection

The coordinate reference system of the subzone layer is checked first.

```{r}
st_crs(mpsz)
```

The tutorial assigns EPSG:3414, the code for Singapore SVY21.

```{r}
mpsz <- st_set_crs(mpsz, 3414)
```

```{r}
st_crs(mpsz)
```

The preschool geometry is inspected before transformation.

```{r}
st_geometry(preschool)
```

```{r}
preschool <- st_transform(preschool, 
                              crs = 3414)
```

```{r}
st_geometry(preschool)
```

The final plot confirms that both layers align.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

### 1.6 Importing and converting aspatial data

The Airbnb listings are ordinary tabular data with longitude and latitude fields.

```{r}
listings <- read_csv("chap01/data/aspatial/listings.csv")
```

```{r}
list(listings) 
```

`st_as_sf()` converts those coordinate columns to point geometry before the result is transformed to SVY21.

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(listings_sf), add = TRUE)
```

### 1.7 Geoprocessing with `sf`

#### Buffering the cycling paths

The cycling paths are first transformed to SVY21 so a buffer can be measured in metres.

```{r}
cyclingpath <- st_transform(
  cyclingpath, crs = 3414)
```

The next cell creates a five-metre buffer around each path.

```{r}
buffer_cycling <- st_buffer(
  cyclingpath, dist=5, nQuadSegs = 30)
```

The buffer area can be added with base R syntax.

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

The tutorial then shows the equivalent `dplyr` approach.

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

```{r}
sum(buffer_cycling$AREA)
```

#### Clipping the buffer to Tampines West

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST") 
```

```{r}
buffer_cycling_selected <- st_intersection(
  buffer_cycling, mpsz_selected)
```

```{r}
plot(st_geometry(buffer_cycling_selected))
```

#### Counting preschools by subzone

`st_intersects()` identifies the points that fall inside each subzone. `lengths()` turns those matches into a count.

```{r}
mpsz$`PreSch Count`<- lengths(st_intersects(mpsz, preschool))
```

```{r}
summary(mpsz$`PreSch Count`)
```

```{r}
top_n(mpsz, 1, `PreSch Count`)
```

Area is needed to compare subzones fairly.

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

```{r}
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

### 1.8 Exploratory data analysis

The first histogram uses base R.

```{r}
hist(mpsz$`PreSch Density`)
```

The next version uses `ggplot2` and adds labels.

```{r}
ggplot(data=mpsz, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning subzones with a single pre-school, on the other hand, \nthere are seven planning subzones with at least 30 or more pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

The scatter plot compares raw counts with density.

```{r}
ggplot(data=mpsz, 
       aes(y = `PreSch Count`, 
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black", 
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```

The chapter page is also exported to the `output/pdf` folder in this project.

```{r}
pdf_dir <- file.path("output", "pdf")
dir.create(pdf_dir, recursive = TRUE, showWarnings = FALSE)
renderthis::to_pdf(from = "https://r4gdsa.netlify.app/chap01.html",
                   to = file.path(pdf_dir, "Hands-on_Ex1a.pdf"))
```

## Part 2: Choropleth mapping

This section follows Chapter 2. It prepares Singapore population data and maps the dependency ratio by planning subzone.

### 2.1 Packages

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

### 2.2 Importing the subzone boundaries

The Master Plan 2019 KML is imported and transformed to SVY21.

```{r}
mpsz <- st_read("chap02/data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_transform(crs = 3414)
```

The KML stores its attributes in HTML. This function retrieves one requested field.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
mpsz
```

### 2.3 Preparing the population table

```{r}
popdata <- read_csv("chap02/data/aspatial/respopagesextod2024.csv")
```

The age groups are combined into young, economically active and aged populations. The dependency ratio compares young and aged residents with the economically active group.

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

The place names are changed to uppercase to match the boundary attributes. Rows without an economically active population are removed.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

### 2.4 Joining population and boundary data

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2024, "chap02/data/rds/mpsz_pop2024.rds")
```

### 2.5 Quick choropleth map

`qtm()` provides a quick first view of the dependency ratio.

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2024, 
    fill = "DEPENDENCY")
```

The tutorial then builds a complete map in one cell.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### 2.6 Building a choropleth step by step

The next five cells add the fill and borders gradually.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders()
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",
             lwd = 0.1,
             lty = "dashed")
```

### 2.7 Comparing classification methods

Quantile classes place a similar number of subzones in each class.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Equal intervals divide the numeric range into classes of the same width.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

Custom breaks can be supplied directly.

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

### 2.8 Choosing colour schemes

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Adding a minus sign reverses the palette.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### 2.9 Adding map elements

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

The legend is positioned automatically inside the map.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

### 2.10 Map styles

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) + 
  tm_borders(fill_alpha = 0.5) + 
  tmap_style("natural")
```

```{r}
tmap_style("white")
```

### 2.11 Small multiple maps

The first method maps the young and aged populations in one `tmap` call.

```{r}
#| fig-width: 12
#| 
tm_shape(mpsz_pop2024) + 
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = 
      tm_legend(position = tm_pos_in(
        "right", "bottom")),
    fill.scale = tm_scale_intervals(
      style = "equal", 
      n = 5,
      values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")

```

The second method creates two maps and arranges them side by side.

```{r}
youngmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                  item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")
                
agedmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
              style = "quantile", 
              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

The dependency ratio can also be faceted by region.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) + 
  tm_facets(by = "REGION_N",
            nrow = 2, 
            ncols = 3,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(fill_alpha = 0.5)
```

### 2.12 Mapping a selected region

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

A box plot is added as a chart inside the map.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

The next map highlights dependency-ratio outliers in red.

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)

outlier_vals <- stats$out

outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile", 
            values = "brewer.blues"),
          fill.legend = tm_legend(),
          fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

### 2.13 Interactive maps

The first interactive map focuses on the Central Region and highlights the same outliers.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
tmap_mode("plot")
```

The final map limits the available zoom levels.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")

tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12,14))
tmap_mode("plot")
```

The Chapter 2 page is saved beside the first PDF in the project's `output/pdf` folder.

```{r}
pdf_dir <- file.path("output", "pdf")
dir.create(pdf_dir, recursive = TRUE, showWarnings = FALSE)
renderthis::to_pdf(from = "https://r4gdsa.netlify.app/chap02.html",
                   to = file.path(pdf_dir, "Hands-on_Ex1b.pdf"))

```

## Conclusion

The two chapters show the full path from raw spatial data to a finished map. The main checks are using a suitable coordinate system, joining records with the correct place name, and choosing a classification method that supports the intended comparison.

## References

- [Chapter 1: Geospatial data wrangling](https://r4gdsa.netlify.app/chap01.html)
- [Chapter 2: Thematic mapping and geovisualisation](https://r4gdsa.netlify.app/chap02.html)

[← Back to Hands-on exercise](hands-on-exercise.qmd){.back-link}
