---
title: "In-Class Exercise 6"
subtitle: "Emerging Hot Spot Analysis"
author: "Le Truong Chinh"
date: "3 October 2026"
date-modified: "4 October 2026"
description: "An ISSS626 in-class exercise tracing how GDP per capita hot and cold spots changed across Hunan from 2005 to 2016."
title-block-banner: true
page-layout: article
toc: true
toc-depth: 4
number-sections: true
code-overflow: wrap
format:
  html:
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
  fig-align: center
---

::: {.exercise-lead}
This exercise looks at where GDP per capita formed hot and cold spots in Hunan, and whether those patterns became stronger or weaker over time.
:::

## Overview

Emerging Hot Spot Analysis combines two ideas. Local Gi\* identifies clusters of high or low values in each year, while the Mann-Kendall test checks whether those clusters show a steady trend through time.

The analysis has four main steps:

1. Create a space-time cube.
2. Calculate local Gi\* for every county and year.
3. Test each county for an upward or downward trend.
4. Classify the resulting hot and cold spot patterns.

## Getting started

### Installing and Loading the R Packages

The packages below handle the spatial data, space-time analysis, maps and interactive chart.

```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```

## The Data

The Hunan boundary layer contains 88 counties. The attribute table records GDP per capita for each county from 2005 to 2016.

### Importing geospatial data

The county boundaries are projected to EPSG 32650 so that distance-based calculations use metres.

```{r}
hunan <- st_read(
  dsn = "data/geospatial", 
  layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

### Importing attribute table

```{r}
GDPPC <- read_csv(
  "data/aspatial/Hunan_GDPPC.csv")
```

## Creating a Time Series Cube

The boundary and yearly GDP records are linked by county name. Each county-year combination becomes one observation in the cube.

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

```{r}
is_spacetime_cube(GDPPC_st)
```

A `TRUE` result confirms that the county and yearly data form a valid space-time cube.

## Computing Gi\*

### Deriving the spatial weights

Counties that share a boundary are treated as neighbours. Each county is also included as its own neighbour because Gi\* measures a county together with its surroundings.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

The same neighbour and weight structure is copied to every year. Row order must remain unchanged after this step.

## Computing Gi\*

Local Gi\* is calculated separately for each year. A positive value points towards a cluster of high GDP per capita, while a negative value points towards a cluster of low GDP per capita.

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

## Mann-Kendall Test

The Mann-Kendall test checks for a consistent trend without assuming that the values follow a normal distribution. Tau is positive for an upward trend and negative for a downward trend. A p-value below 0.05 indicates a statistically significant trend.

### Mann-Kendall Test on Gi

Changsha is used as a simple example before the test is repeated for all counties.

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

```{r}
#| eval: false
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

```{r}
#| echo: false
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

### Interactive Mann-Kendall Plot

The same series can be explored interactively by moving over the line.

```{r}
#| eval: false
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

```{r}
#| echo: false
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

### Printing Mann-Kendall test report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

Here, `tau` gives the direction and strength of the trend, while `sl` is the p-value used to judge its significance.

### Mann-Kendall test data.frame

The same test is now applied to every county.

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```

### Mann-Kendall test data.frame

The table below brings the strongest and most statistically significant trends to the top.

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```

## Performing Emerging Hotspot Analysis

The full analysis combines the yearly Gi\* results with each county's trend. One time lag is used, with 99 simulations for the significance test.

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

### Visualising the distribution of EHSA classes

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()

```

Sporadic cold spots are the largest group, followed by sporadic hot spots. This means most detected counties move in and out of significance instead of remaining a hot or cold spot throughout the whole period.

### Visualising EHSA

The EHSA results are joined to the county boundaries. Only classes with a p-value below 0.05 are coloured on the final map.

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))

```

```{r}
#| eval: false
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```

```{r}
#| echo: false
#| fig-height: 10
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```

The significant classes are spread across Hunan rather than forming one single part of the province. The map also supports the bar chart: sporadic patterns are much more visible than long-lasting ones.

### Interpretation of EHSA classes

- **Persistent:** the county remained a statistically significant hot or cold spot for most of the period.
- **Intensifying:** the spatial pattern became stronger over time.
- **Diminishing:** the pattern remained present but became weaker.
- **New:** the county became a hot or cold spot only near the end of the period.
- **Sporadic:** the pattern appeared in some years but was not continuous.

## Takeaways

Looking at one year alone would miss how the spatial pattern developed. Gi\* identifies where neighbouring counties share high or low GDP per capita in each year, while the Mann-Kendall test distinguishes a short-lived pattern from a consistent change through time.

## References

- [Chapter 11: Emerging Hot Spot Analysis](https://r4gdsa.netlify.app/chap11)
- [Understanding Emerging Hotspots with sfdep](https://sfdep.josiahparry.com/articles/understanding-emerging-hotspots.html)
