---
title: "In-Class Exercise 4"
author: "Le Truong Chinh"
format: html
date: "September 18, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---
### Loading the package
To complete this exercise, sf, ggstatsplot, tmap, tidyverse, knitr and GWmodel will be used.
```{r}
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)
```

### Preparing the data
### Import shapefile into r environment
```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```
### Import csv file into r environment

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```
### Joining Hunan and Hunan_2012
```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:3, 7, 15, 16, 31, 32)
```
### Mapping GDPPC
```{r fig.width=16, fig.height=6}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```
## Converting to SpatialPolygonDataFrame
```{r}
hunan_sp <- hunan %>%
as_Spatial()
```
## Geographically Weighted Summary Statistics (GSWW) with adaptive bandwith
```{r}
bw_CV <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```
```{r}
bw_AIC <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "AICC",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```
```{r}
bw_CV
```
```{r}
bw_AIC
```
### Computing geographically weighted summary statistics
```{r}
gwstat <-gwss(data = hunan_sp,
              vars = "GDPPC",
              bw = bw_AIC,
              kernel = "bisquare",
              quantile = TRUE,
              adaptive = TRUE,
              longlat = T)
```

### Preparing the output data
```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
hunan_gstat <- cbind(hunan, gwstat_df)
```
### Visualising geographically weighted summary statistics
```{r fig.width=16, fig.height=6}
tm_shape(hunan_gstat) + 
  tm_polygons(fill="GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha=0.5) + 
  tm_title("Distribution of geographically weighted mean") +
  tm_layout(legend.position = c("left", "bottom"))
```
